---
title: "Data Visualisation Assignment 2"
author: "Saurabh"
date: "2025-09-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



```{r}
getwd()
```



# Exploring IMDb dataset


# Part A: Data pre-processing

```{r}
library(tidyverse)
library(lubridate)   
library(janitor)     
library(stringr)     

# 1. Import the datasets

movies_raw <- read_csv("../data/movies.csv")
budget_raw <- read_csv("../data/budget.csv")

movies_raw
budget_raw

```


```{r}
# 2. Clean names 

movies <- movies_raw %>% clean_names()
budget <- budget_raw %>% clean_names()

movies
budget

```




```{r}

# 3. Convert release_date style to proper date format(YYYY-MM-DD)

budget <- budget %>%
  mutate(release_date = mdy(release_date) )


# 4. Remove missing dates

budget <- budget %>%
  filter(!is.na(release_date))

budget
```

# Investigate and explain why certain dates failed to convert.

While converting the release_date column with the mdy() function, two rows couldn’t 
be turned into proper dates. 
A closer look showed that these records listed only a four-digit year (like “2025” and “2027”)
instead of a full month-day-year date such as “Jun 1, 2011”. 
Because mdy() requires a complete date to work, it left those entries as NA. 
In line with Step 4 of the task, we removed these rows so that the dataset keeps 
only records with fully specified, valid release dates.


```{r}


# 5. Create year column

budget <- budget %>%
  mutate(year = year(release_date))

budget

# 6. Inner join 

movies <- movies %>%
  mutate(primary_title = str_to_lower(primary_title))

budget <- budget %>%
  mutate(movie = str_to_lower(movie))



movies_budget <- inner_join(movies, budget, 
                            by = c("primary_title" = "movie", "start_year" = "year"))

movies_budget


nrow(budget)
nrow(movies_budget)

```

# 7. Explain why the resulting dataset has fewer rows than the original budget dataset.

The final joined dataset is noticeably smaller because many rows from the budget 
table could not find a perfect match in the movies table.
Two key factors explain this:


1. A few rows were dropped during cleaning.
   Some ReleaseDate values in the budget data contained only a four-digit year, 
   such as “2025” or “2027”.
   The mdy() function requires a complete month–day–year date, so these entries 
   could not be converted and were removed when rows with missing dates were filtered out.
   This accounts for only a small portion of the reduction.
   

2. Most of the reduction is due to strict join requirements.
   The inner_join() keeps only records where the lower-cased movie title and 
   release year match exactly in both datasets.Many budget rows don’t meet this 
   standard because of small but critical differences for example, punctuation or
   spacing changes, alternate or shortened titles, the use of “&” instead of “and”, 
   or the presence of colons, dashes, or parenthetical notes such as “Director’s Cut”.
   Lowercasing helps with letter case but does not fix these deeper textual mismatches.
   In other cases the release year differs slightly, as IMDb may record a festival
   premiere or an international release year that does not align with the calendar 
   year in the budget data. Rows with missing years on either side also cannot join.
   
   
Because an inner join only retains rows that match on both movie title and release
year, all entries with these kinds of inconsistencies are excluded.
As a result, the final movies_budget table contains about 8,388 rows, compared 
with 15,286 rows in the cleaned budget data.
   


# Part B: Data visualisation


# Day of the week for movie releases
```{r}
# Add a weekday column using wday() and label the days
movies_budget <- movies_budget %>%
  mutate(release_wday = wday(release_date, label = TRUE, abbr = FALSE))

# Visualise the count of movies by release weekday
ggplot(movies_budget, aes(x = release_wday)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Movie releases by day of the week",
       x = "Day of the Week",
       y = "Number of Movies") +
  theme_minimal()

```

# On which days are movies most frequently released?
Movies were released most frequently on friday.

# Explain what this suggests about release strategies in the film industry.
Most movies are released on friday because it is the start of the weekend. 
As majority of people don't go to work on saturdays and sunday, so many people 
spend their free time to watch movies in theatre starting from Friday evening.




# Box office performance by release day
```{r}
revenue_by_wday <- movies_budget %>%
  mutate(release_wday = wday(release_date, label = TRUE, abbr = FALSE)) %>%
  group_by(release_wday) %>%
  summarise(mean_domestic = mean(domestic, na.rm = TRUE), .groups = "drop") %>%
  mutate(release_wday = fct_relevel(release_wday,
                     "Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))



ggplot(revenue_by_wday, aes(x = release_wday, y = mean_domestic)) +
  geom_col(fill = "skyblue", width = 0.4) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +  # show millions
  labs(title = "Average domestic revenue by release day",
       x = "Day of the Week",
       y = "Average Domestic Revenue (Millions USD)") +
  theme_minimal()

```


# Do movies released on certain days perform better at the domestic box office revenue?
The movies which released on Thursday performed better and generated
better revenue.


# Interpret the patterns observed in your visualisation.
The chart shows that movies released on Thursdays earn the highest average 
domestic revenue of about 62 million dollars, followed closely by Wednesdays at
around 55 million dollars and Tuesdays at about 48 million dollars. 
Fridays still perform well at roughly 40 million dollars, but they trail the 
mid week releases. Mondays average around 30 million dollars, while Sundays and 
Saturdays bring in the least, at about 25 million and 20 million dollars 
respectively.
The analysis shows that films opening on Thursdays record the highest average 
domestic gross followed by Wednesdays and Tuesdays, while Saturday releases earn
the least. This suggests that the day a film starts its run can influence its 
total domestic box office performance.



# Domestic box office trends by month
```{r}
# Extract month and group by it
movies_budget <- movies_budget %>%
  mutate(release_month = month(release_date, label = TRUE, abbr = TRUE))

# Summarise total domestic revenue by month
revenue_by_month <- movies_budget %>%
  group_by(release_month) %>%
  summarise(total_domestic = sum(domestic, na.rm = TRUE))


library(scales)

ggplot(revenue_by_month, aes(x = release_month, y = total_domestic, group = 1)) +
  geom_line(color = "forestgreen", size = 0.5) +
  geom_point(color = "forestgreen", size = 2) +
  scale_y_continuous(labels = label_number(scale = 1e-9, suffix = "B")) +  # show in billions
  labs(title = "Domestic box office trends by month",
       x = "Month of Release",
       y = "Total Domestic Revenue (Billion USD)") +
  theme_minimal()

```


# Are there particular months in which movies tend to earn higher domestic revenue?
Movies released in June, July, November and December earn the higher domestic revenue.


# Discuss factors that might influence these trends.
The higher revenues in June, July, November and December are largely driven by 
school holidays, summer vacations and major festive breaks like Thanksgiving 
and Christmas. These periods give families and students more free time and 
increase leisure activities, making people more likely to visit cinemas. 
Studios also plan big budget releases during these months to tap into the larger
audiences and maximize earnings.


# Worldwide revenue over time by genre
```{r}
# Summarise yearly worldwide revenue per genre
worldwide_by_genre <- movies_budget %>%
  group_by(start_year, genre) %>%
  summarise(total_worldwide = sum(worldwide, na.rm = TRUE), .groups = "drop")


ggplot(worldwide_by_genre, aes(x = start_year, y = total_worldwide, fill = genre)) +
  geom_area(alpha = 0.8 , size = 0.3, colour = "white") +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = "B")) +
  labs(
    title = "Worldwide box office revenue over time by genre (stacked)",
    x = "Release Year",
    y = "Total Worldwide Revenue (USD billions)",
    fill = "Genre") +
  theme_minimal()

```


# Interpretation

Action (red area):
Action movies have been the main driver of global box office success. 
Their earnings rose strongly from the 1990s, peaking around 2015–2018. 
Even though there’s been a slight dip in the last few years, action films still 
bring in the highest worldwide revenue.

Adventure (yellow area):
Adventure films have grown steadily alongside action since the 1990s, 
reaching their highest point in the mid-2010s. They remain a close second to 
action movies in overall box office performance.

Comedy (green area):
Comedies were strong performers through the 1990s and early 2000s but have 
slowed down since about 2010. One reason could be that humor doesn’t always 
translate well across different cultures, making these films less popular 
globally.

Drama (blue area):
Drama movies grew steadily until the early 2000s but have gradually declined 
since then. While they often receive critical praise and win awards, they
generally don’t attract as big an audience as action or adventure films.

Horror (purple area):
Horror films have always been smaller in total revenue, though they’ve shown 
some growth in recent years. Because they’re usually cheaper to make, horror 
films can still be very profitable even with lower worldwide earnings.

Unclassified/NA (grey area):
Movies without a specific genre label have stayed very small over time and
haven’t had much effect on worldwide box office trends.



The sharp drop around 2020 aligns with pandemic-related cinema closures and 
the rebound afterwards reflects theatres reopening and blockbuster releases 
returning. Overall, global audience preferences and large scale franchise 
productions have shifted the revenue balance strongly toward action oriented 
genres.


# Return on Investment (ROI)
```{r}
movies_budget <- movies_budget %>%
  mutate(roi = worldwide / production_budget)

# ROI vs production budget
ggplot(movies_budget, aes(x = production_budget, y = roi, colour = genre)) +
  geom_point(alpha = 0.4) +
  scale_x_continuous(labels = scales::label_dollar(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(
    title = "ROI vs Production Budget by Genre",
    x = "Production Budget (USD millions)",
    y = "ROI (Worldwide / Production Budget)"
  ) +
  theme_minimal()

```





```{r}
# ROI vs worldwide revenue
ggplot(movies_budget, aes(x = worldwide, y = roi, colour = genre)) +
  geom_point(alpha = 0.4) +
  scale_x_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = "B")) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(
    title = "ROI vs Worldwide Gross by Genre",
    x = "Worldwide Revenue (USD billions)",
    y = "ROI"
  ) +
  theme_minimal()

```


# Interpretation

# ROI vs Worldwide gross by Genre

Horror (purple):
Horror films often deliver striking returns. Many are made on very small budgets
yet manage to generate huge profits when they connect with audiences. These big 
wins explain why horror has a reputation for low cost but high reward.

Action (red):
Action movies dominate worldwide revenue but their ROI usually remains modest. 
Huge budgets and large marketing spends mean that even billion-dollar box office
totals often translate into relatively lower ROI.

Adventure (olive):
Adventure films also generate strong global revenue, often close to action films, 
but with ROI generally in the same modest range. Big production and marketing 
costs limit overall return compared to the headline revenue numbers.

Comedy (green):
Comedies mostly fall in the mid-range of ROI. They can do well in certain markets,
but humour doesn’t always travel easily across cultures, making worldwide earnings
less predictable and limiting the potential for very high ROI.

Drama (blue):
Drama films show mixed and scattered ROI. While some achieve critical acclaim 
and occasional box office success, they typically don’t generate the massive 
returns of action, adventure, or even surprise-hit horror films.


# ROI vs Budget by Genre

Horror:
Typically produced with very small budgets, horror occasionally delivers runaway
successes with ROI well above 100. Although risky, when a film hits, it can be 
hugely profitable.

Action:
Action films are concentrated at very large budgets and usuallyachieve ROI between 
1 and 10. They dominate the global box office but require
heavy investment to reach those earnings.

Adventure:
Adventure movies also lean toward high budgets and show ROI generally similar
to action films. They benefit from broad appeal but need major financial 
backing to succeed.

Comedy:
Comedies often operate with mid-sized budgets and have highly variable ROI.
A few standouts perform very well, but many struggle to produce consistent 
worldwide returns.

Drama:
Drama films also work with mid-sized budgets but show even more variability in ROI.
While some titles earn strong praise or niche success, most lack the global draw 
for sustained high ROI.

Unclassified (NA):
These films show no consistent pattern and make little impact on overall trends.



# Top production companies
```{r}
library(forcats)

# Lump less frequent companies into "Other"
top_companies <- movies_budget %>%
  mutate(prod_co_grouped = fct_lump_n(production_company, n = 10)) %>%
  group_by(prod_co_grouped, genre) %>%
  summarise(total_worldwide = sum(worldwide, na.rm = TRUE), .groups = "drop")


ggplot(top_companies,
       aes(x = reorder(prod_co_grouped, total_worldwide),
           y = total_worldwide,
           fill = genre)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = "B")) +
  labs(
    title = "Top production companies by worldwide revenue (genre)",
    x = "Production Company",
    y = "Total Worldwide Revenue (USD billions)",
    fill = "Genre"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, 
                              margin = margin(b = 15)),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20))


```

# Interpretation

# Top production companies by worldwide revenue and genre

Action:
Universal Pictures and Warner Bros. are the clear leaders in action films, 
regularly turning out big-budget blockbusters that dominate worldwide box offices.
Columbia Pictures and Paramount also make strong contributions, but their totals
remain behind the top two.

Adventure:
Walt Disney Pictures stands out in adventure films, thanks to global hits like 
Marvel, Star Wars and Pixar series. Universal Pictures and Warner Bros. follow 
with steady earnings, but Disney’s worldwide reach and strong franchises give it
a clear edge.

Comedy:
Universal Pictures and Warner Bros. are also strong in comedy, producing a steady
stream of crowd-pleasing hits. Columbia Pictures adds solid support, known for 
family-friendly and romantic comedies with broad appeal.

Drama:
Paramount Pictures and Columbia Pictures lead in drama, backing many award winning 
and popular titles. Universal and Warner Bros. contribute as well but with slightly
lower overall drama revenue.

Horror:
New Line Cinema and Lionsgate shine in horror, focusing on low cost and high return 
productions. Their ability to make profitable movies on smaller budgets lets them 
compete effectively with the bigger studios.

Unclassified/NA:
The "Other" group is a mix of many smaller and independent studios, together 
generates the largest single share overall. While each studio is small on its own,
their combined output across genres makes a major impact worldwide.





















